\begin{abstract}
Most perturbation methods ask what an intervention will do. A complementary question comes
first: can the available intervention class even point the transcriptome in the desired
direction? We introduce a convex-cone feasibility test that treats a genome-scale CRISPRi
Perturb-seq screen as a dictionary of measured effect vectors. Under an explicit additive
approximation, non-negative least squares projects a target direction onto combinations of
knockdown effects whose orientation cannot be reversed. The projection yields a ranked
mixture and a greedy sparse candidate panel; a non-zero residual yields a Farkas/KKT dual
certificate that proves the full target direction lies outside the measured knockdown cone.
Positive residual coordinates rank unmet readouts for CRISPRa or de-repression follow-up, but
do not by themselves validate those interventions. In primary human
CD4\textsuperscript{+} T cells, the T\textsubscript{H}2$\rightarrow$T\textsubscript{H}1
direction is only partly captured by knockdown (fixed-split held-out cosine $0.448$, larger
than all $60$ shuffled targets; empirical $p=1/61$). A staged, sign-symmetry proxy assigns
$39\%$ of target energy to measured loss-of-function directions, $25\%$ to sign-flipped
gain-of-function directions, and $35\%$ to neither. Across 12 target--condition cases,
loss-of-function is never the majority component (mean $0.34$), although the eight-shuffle
atlas nulls are screening estimates rather than precise tail probabilities. The same operator
runs without retuning on a held-out CEBPA state in a K562 CRISPRa screen (cosine $0.878$);
$126$ measured doubles show that additivity is bounded (median cosine $0.71$), not exact. A
structured survey of $91$ prior methods found no prior method combining measured-effect
grounding with this target-specific cone certificate. We therefore position the method as a
directional experiment-triage instrument, not a proof of phenotypic rescue or target validity.
\end{abstract}
